Software Clusterings with Vector Semantics and the Call Graph

### 2019-04-26 manuscript/manuscript.tex
Move manuscript under manuscript/ directory

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Proposing Software Clusterings using Vector Semantics and the Call Graph: A Study on the Linux Kernel}

@@

\noindent \textbf{Purpose -- } Propose a method to determine a software's modules without 
knowledge of its structure, and empirically validate the method's performance.

@@

Modern software systems usually span hundreds of thousands of source code lines (SLOC) 
since the inherent complexity of modern applications that computers have to solve has 
exponentially increased. The need for adoption of modular software architectural schemas 
becomes inevident for robust maintainability of the overall software system.
Since the architecture of a software system is the most fundamental realization of a 
software system, when there is no description of the architecture of it, software 
architects usually make attempts to recover it. In the past, there have been attempts to 
identify the main components of a software system using hierarchical clustering 
algorithms \cite{maqbool_overview} mainly by clustering modules into subsystems, given 
its source code.

Plenty of attention has been gathering around the variety of possibilities enabled by 
modern machine learning techniques and practices in Natural Language Processing (NLP) 
field and bought to the forefront by numerous open-source NLP projects, such as spaCy 
\cite{spacy} and NLTK \cite{nltk}, performing analysis on large corpuses of data. These 
practices can also be applied on codebases as corpuses of text, thus enabling researchers 
to extract useful information for the linguistic contents of the source codebase 
inaugurating new modes of reviewing large amounts of source code. So far, NLP practices 
have been a subject to systematic research for codebases, one attempt of them being the 
Deep Code Search project \cite{deepcodesearch}.

This paper aims to propose software clusterings using a combination of vector semantics on the source code provided by \emph{Doc2Vec} as well as the existing structure of 
the codebase via the call graph in order to construct an augmented call graph and form communities in this graph ariving at a module--level 
architecture viewed from the call graph perspective. Our approach will be applied on the 
Linux Kernel codebase and will be compared to its existing layered architecture using 
MoJo distance as a clusterings metric, proposed by Tzerpos and Holt in reference \cite{mojo}, 
as well as, compare its stability and extremity compared to other Clustering 
Algorithms that exist in literature.

@@

@@

For the preprocessing stage use the NLP package spaCy \cite{spacy} which provided us with very useful tools for the extraction. After preprocessing, we use gensim \cite{gensim} to train the Doc2Vec model. First of all, we remove the stopwords that occured in the text like control flow keywords like \texttt{if}, semicolons, variable types like \texttt{int} and so on. Next, we tokenize each word splitting it to its constituent parts regarding its form. For example the method \texttt{\_\_zone\_seqlock\_init} corresponds to \texttt{zone}, \texttt{seqlock} and \texttt{init}. Then the identifiers are split using a dynamic programming approach, so the identifier \texttt{inprogress} is split into \texttt{in} and \texttt{progress}. Then we pass the tokens from an English lemmatizer to remove excess information for words. For example a token for an array named \texttt{literals} becomes \texttt{literal}. After that, the vector size for the Doc2Vec algorithm is set according to the size of the codebase which is being examined each time, in the order of magnitude of a few hundreds since for smaller datasets with poorer vocabularies, a lower dimension
should suffice for capturing the necessary features.  Finally, the document tags corresponded to the source code files or modules and the algorithm are assigned a vector $\vec u_v \in \mathbb R^d$ for each $v \in V$.

@@

\paragraph{Combining semantics and the Call Graph} The input to the clustering procedure
 is the call graph with cosine similarities of the various modules as weights, thus forming a
  weighted directed graph. Then we run the Louvain Clustering Algorithm \cite{louvain} in
   order to obtain the software clustering via maximizing the modularity function with a
    greedy approach. Further refining can also be done by re-running the algorithm on the contracted graph with mean embeddings as embeddings for each node.

\paragraph{Implementation}  Our method is implemented in the Python Language using spaCy\cite{spacy} for the 
NLP tasks, Gensim \cite{gensim} for the generation of Document Embeddings. The library 
used to handle graphs and graph operations was NetworkX \cite{nx}. The code is available on GitHub.

\paragraph{Decision Rationale for Evaluation System} We have chosen to use the Linux Kernel as a codebase to evaluate our method on since it is a large and complex system with 27 years of continuous development, spanning 20.3 millions SLOC at the time of writing.  Moreover, it is easy to find ground truth system due to its clear directory structure and construct testcases for
evaluations. Finally, it was also used in related studies \cite{acdc, evaluation} as a reference system for evaluation.

\paragraph{Experimental Setting} Our method was run on the Linux Kernel 4.21,
 consisting of 20.3 million SLOC against Average-Linkage \cite{average}, Complete-Linkage \cite{complete} and Ward-Linkage \cite{ward} as well as ACDC and LIMBO. The results are presented in Table \ref{table:evaluation}. As ground truth we have used the first level directories as a target clustering and as input we have considered the modules of the one-top directories. For example, the source code file \texttt{drivers/net/ieee802154/mcr20a.c} has a ground truth value of \texttt{drivers} and it is considered under the same module as every \texttt{.c} and \texttt{.h} file under \texttt{drivers/net/ieee802154}. As a clustering distance metric we have used the MoJo metric, as proposed in reference \cite{mojo}. Since, Louvain Community Detection produces 
 different (but very similar) results every time it runs, we ran the simulations multiple times and averaged the results.

\section{Discussion of Results}

@@

Firstly, our approach produced balanced clusterings thus demonstrating low \emph{extremity} compared to the other clusterings, producing an amount of clusters that was close to the Ground Truth with smaller standard deviation in the cluster sizes than the other clustering methods, without knowing the number of clusters \emph{a priori}. Besides this, our approach also seems to give balanced clusters with respect to \emph{Median Cluster} criterion. Note that ACDC, which took into account only the structural properties of the files, produced a high granularity clustering that failed to catch the structure of the Linux Kernel Codebase.

Moreover, in the \emph{authoritativeness} criterion, our approach outperformed the other clusterings to a very up to 10 times in terms of MoJo distance being very close (smaller is better) to the Ground Truth, with a very small standard deviation. This also suggests great
\emph{stability} in the resulting clusterings. Note that the other clustering methods we compared against ours gave disappointing MoJo distances even with their cluster number defined 
equal to the one of the Ground Truth.

@@

% Architectural recovery is an active research 
%field aiming to make software systems more understandable and maintainable over time. The 
%study of architectural change requires that the architecture at a given point in time, 
%for example after a major release, during a system's evolution must be inferred. 
Various algorithms have been proposed for realizing architectural schemata many of which use 
different features for describing the source code.

Onaiza Maqbool and Haroon A. Babri \cite{maqbool_overview} present an
overview of the various approaches taken towards to hierarchical clustering algorithms
for software systems. In their study, they compare the various clustering algorithms
such as ACDC, LIMBO and other Traditional Clustering Algorithms such as Single
Linkage, Complete Linkage \cite{complete} and Weighted Linkage presenting their efficiency regarding 
multiple feature sets.

Added to this, a \emph{semantics-based architectural view} of the system, as discussed in reference
\cite{large_study} reveals significant aspects of a software system and its change over 
time, which suggests that semantic-based approached should be followed for gaining better 
understandings of a software system. Their suggestion is a \emph{term frequency--inverse 
document frequency} (TF-IDF) model serving as a similarity metric between source code 
entities.

Work in Software Clustering Algorithms (SCA) has been extensively done by Tzerpos in
references \cite{MoJoFM}, \cite{stability}, \cite{acdc} and \cite{limbo} with the ACDC
and LIMBO clustering algorithms.

In particular, the ACDC algorithm leans toward to software components comprehension 
based on subsystem patterns. Their approach considers an initial structure of the system, without taking into account semantics, and tries to build comprehensive clusterings of the given ground structure. The authors also conducted a study on an older version of  the Linux Kernel.

Finally, Andritsos and Tzerpos in research article \cite{limbo} present an Information--Theoretic 
approach of SCA by developing LIMBO which clusters modules upon inserting their Distributional Cluster Features to a B-tree variant and then applying the \emph{Agglomerative Information Bottleneck} algorithm, merging the clusters with the minimum mutual information loss together.

\section{Conclusions} This paper is set out to show how we can use a combination of vector semantics
and information from the call graph in order to produce meaningful clusterings of a software system.

Taking into consideration the trends and trajectories of modern software architecture recovery techniques, we have conducted our study in a huge codebase spanning 20.3 millions of SLOC and we have outperformed
several clustering algorithms of the existing literature in terms of \emph{extremity} and \emph{authoritativeness} as well as we have produced \emph{stable} clusterings of the codebase 
without even the knowledge of the number of clusters of the Ground Truth. This evidence strongly
supports the further usage of vector semantics and structural information of the project for the 
recovery of the underlying architecture of a software system.

Within the context of this project further work is encouraged in several areas such identifier splitting, 
which can be done using deep learning \cite{splitting}, integration with more static analyzers and usage 
of the project in order to propose layered architectures by finding tree structures on the resulting clustering. 
Finally, it is  encouraged to try other methods described thoroughly in \cite{malliaros}, such as using a 
different modularity function or using spectral clustering and cut-based measures.

\begin{table*}
  \caption{Experimental Results for the Linux Kernel 4.21 Codebase}
    \label{tab:evaluation}
    \begin{tabular}{lrrrrrr}
    \toprule
    Algorithm & Number of Clusters & Cluster Size Range & Avg. Cluster & Standard Deviation & Median Cluster & MoJo Distance \\
    \midrule
    ACDC \cite{acdc} & 9055 & 1 -- 4245 & 4.53 & 45.82 & 2\\
    Average Linkage \cite{average} & 21 & 1--3406 & 163 & 725.11 & 1 & 2092 \\
    Complete Linkage \cite{complete} & 21 & 1--1529 & 163 & 412.34 & 19.0 & 1710 \\
    LIMBO \cite{limbo} & & \\
    SADE & 10 (\pm 2)  & 2 (\pm 0) -132 (\pm 13) & 64 (\pm 4) & 40 (\pm 4) & 65 (\pm 10) & 243 (\pm 1)  \\
    SADE (Directed) & 5 (\pm 2) & 1 (\pm 1) - 614 (\pm 1) & 141 (\pm 39) & 253 (\pm 25) & 2 (\pm 0.3)  & 237 (\pm 2) \\

\newpage
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

>>>>

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title{Software Clusterings with Vector Semantics and 
the Call Graph}

@@

\noindent \textbf{Purpose -- } Propose a method to determine a software's modules without 
knowledge of its architectural structure, and empirically validate the method's performance.

@@

%Modern software systems usually span hundreds of thousands of source code lines (SLOC) 
%since the inherent complexity of modern applications that computers have to solve has 
%exponentially increased.

In modern software systems, the need for adoption of modular software architectural schemas 
becomes inevident for robust maintainability of the overall software system.
Since the architecture of a software system is the most fundamental realization of a 
software system, when there is no description of the architecture of it, software 
architects usually make attempts to recover it. In the past, there have been attempts to 
identify the main components of a software system using hierarchical clustering 
algorithms \cite{maqbool_overview, limbo} mainly by clustering modules into subsystems, given 
its source code.    
The possibilities offered by machine learning techniques and practices in the Natural Language Processing (NLP) 
can also be applied on codebases as corpuses of text, thus enabling 
the extraction of useful information for the semantics of the source codebase 
inaugurating new modes of reviewing large amounts of source code.
An example is demonstrated through the Deep Code Search Project \cite{deepcodesearch} .

This paper aims to propose software clusterings using a combination of vector semantics on the source code provided by \emph{Doc2Vec} as well as the existing structure of 
the codebase via the call graph in order to form communities adn arrive at a module-level 
architecture viewed from the call graph perspective. 
We will study Linux codebase and we will compare it with ground truth in using the MoJo metric \cite{mojo} as well as 
compare it with other state-of-the-art and baseline clustering algorithms \cite{maqbool_overview, evaluation} in terms of 
\emph{stability}, \emph{authoritativeness} and \emph{extremity}

@@

\begin{table*}
  \caption{Experimental Results for Linux 4.21 Codebase}
    \label{tab:evaluation}
    \begin{tabular}{lrrrrrr}
    \toprule
    Algorithm & Number of Clusters & Cluster Size Range & Avg. Cluster & Standard Deviation & Median Cluster & MoJo Distance \\
    \midrule
    ACDC \cite{acdc} & 9055 & 1 -- 4245 & 4.53 & 45.82 & 2\\
    Average Linkage \cite{average} & 21 & 1--3406 & 163 & 725.11 & 1 & 2092 \\
    Complete Linkage \cite{complete} & 21 & 1--1529 & 163 & 412.34 & 19.0 & 1710 \\
    LIMBO \cite{limbo} & & \\
    SADE & 10 (\pm 2)  & 2 (\pm 0) -132 (\pm 13) & 64 (\pm 4) & 40 (\pm 4) & 65 (\pm 10) & 243 (\pm 1)  \\
    SADE (Directed) & 5 (\pm 2) & 1 (\pm 1) - 614 (\pm 1) & 141 (\pm 39) & 253 (\pm 25) & 2 (\pm 0.3)  & 237 (\pm 2) \\

@@

For the preprocessing stage use the NLP package spaCy \cite{spacy} which provided us with very useful tools for the extraction. After preprocessing, we use gensim \cite{gensim} to train the Doc2Vec model. First of all, we tokenize the code and remove the stopwords. After that, for each token, we split it into its constituent parts using dynamic programming and lemmatize 
each individual sub-token. 
For example the method \texttt{\_\_zone\_seqlock\_init} corresponds to \texttt{zone}, \texttt{seqlock} and \texttt{init}, \texttt{inprogress} is split into \texttt{in} and \texttt{progress} and \texttt{literals} becomes \texttt{literal}.

@@

\paragraph{Combining semantics and the Call Graph} The input to the clustering procedure
 is the call graph with cosine similarities of the various modules as weights, thus forming a
  weighted directed graph. Then we run the Louvain Clustering Algorithm \cite{louvain} in
   order to obtain the software clustering via maximizing the modularity function with a
    greedy approach.

    %Further refining can also be done by re-running the algorithm on the contracted graph with mean embeddings as embeddings for each node.

\paragraph{Implementation}  Our method is implemented in the Python Language using spaCy\cite{spacy} Gensim \cite{gensim} and NetworkX \cite{nx}.

% used to handle graphs and graph operations was NetworkX \cite{nx}. The code is available on GitHub.

\paragraph{Decision Rationale for Evaluation System} We have chosen to use Linux as a codebase to evaluate our method on since it is a large and complex system with 27 years of continuous development, spanning 20.3 millions SLOC at the time of writing.  Moreover, it is easy to find ground truth system due to its clear directory structure and construct testcases for
evaluations. Finally, it was also used in related studies \cite{acdc, evaluation} as a reference system for evaluation.

\paragraph{Experimental Setting} Our method was run on Linux 4.21,
 consisting of 20.3 million SLOC against Average-Linkage \cite{average}, Complete-Linkage \cite{complete} and Ward-Linkage \cite{ward} as well as ACDC and LIMBO. The results are presented in Table \ref{tab:evaluation}. As ground truth we have used the first level directories as a target clustering and as input we have considered the modules of the one-top directories. For example, the source code file \texttt{drivers/net/ieee802154/mcr20a.c} has a ground truth value of \texttt{drivers} and it is considered under the same module as every \texttt{.c} and \texttt{.h} file under \texttt{drivers/net/ieee802154}. As a clustering distance metric we have used the MoJo metric, as proposed in reference \cite{mojo}. Since, Louvain Community Detection produces 
 different (but very similar) results every time it runs, we ran the simulations multiple times and averaged the results.

\section{Results, Evaluation, Discussion}

@@

Firstly, our approach produced balanced clusterings thus demonstrating low \emph{extremity} compared to the other clusterings, producing an amount of clusters that was close to the Ground Truth with smaller standard deviation in the cluster sizes than the other clustering methods, without knowing the number of clusters \emph{a priori}. Besides this, our approach also seems to give balanced clusters with respect to \emph{Median Cluster} criterion. Note that ACDC, which took into account only the structural properties of the files, produced a high granularity clustering that failed to catch the structure of Linux Codebase. 
Moreover, in the \emph{authoritativeness} criterion, our approach outperformed the other clusterings to a very up to 10 times in terms of MoJo distance being very close (smaller is better) to the Ground Truth, with a very small standard deviation. This also suggests great
\emph{stability} in the resulting clusterings. Note that the other clustering methods we compared against ours gave disappointing MoJo distances even with their cluster number defined 
equal to the one of the Ground Truth.

@@

% Architectural recovery is an active research 
%field aiming to make software systems more understandable and maintainable over time. The 
%study of architectural change requires that the architecture at a given point in time, 
%for example after a major release, during a system's evolution must be inferred. 
%Various algorithms have been proposed for realizing architectural schemata many of which use 
%different features for describing the source code.

Onaiza Maqbool and Haroon A. Babri \cite{maqbool_overview} present an
overview of the various approaches taken towards to hierarchical clustering algorithms
for software systems. In their study, they compare the various clustering algorithms
such as ACDC, LIMBO and other Traditional Clustering Algorithms such as Single
Linkage, Complete Linkage \cite{complete} and Weighted Linkage presenting their efficiency regarding 
multiple feature sets. 
Added to this, a \emph{semantics-based architectural view} of the system, as discussed in reference
\cite{large_study} reveals significant aspects of a software system and its change over 
time, which suggests that semantic-based approached should be followed for gaining better 
understandings of a software system. Their suggestion is a \emph{term frequency--inverse 
document frequency} (TF-IDF) model serving as a similarity metric between source code 
entities. 
Work in Software Clustering Algorithms (SCA) has been extensively done by Tzerpos in
references \cite{mojo}, \cite{stability}, \cite{acdc} and \cite{limbo} with the ACDC
and LIMBO clustering algorithms.

In particular, the ACDC algorithm leans toward to software components comprehension 
based on subsystem patterns. Their approach considers an initial structure of the system, without taking into account semantics, and tries to build comprehensive clusterings of the given ground structure. The authors also conducted a study on an older version of  Linux. 
Moreover, Andritsos and Tzerpos in reference \cite{limbo} present an Information--Theoretic 
approach of SCA by developing LIMBO which clusters modules upon inserting their Distributional Cluster Features to a B+-tree variant and then applying the \emph{Agglomerative Information Bottleneck} algorithm, merging the clusters with the minimum mutual information loss together.

\section{Conclusions} This paper is set out to show how we can use a combination of vector semantics
and information from the call graph in order to produce meaningful clusterings of a software system.
We conclude that we have outperformed state-of-the-art software clusterings and conventional
agglomerative clustering algorithms 
in terms of \emph{extremity}, \emph{authoritativeness} and \emph{stability} 
without knowing the number of clusters of the Ground Truth. This evidence strongly
supports the further usage of vector semantics and the call graph for architecture recovery.
Finally, further work is encouraged in several areas such as 
integration with more static analyzers and use of the project in order to propose layered architectures.

\newpage
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}
